Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-12-10 rnni_convexity.tex
ne need shorter sudo-codes
hope for thm:caterpillar_convex

\begin{algorithm}[H]
\caption{$\findpath$($T,R$)}
\label{alg:find_path}
\begin{algorithmic}[1]
\STATE $T' := T$
\STATE $p := [T']$
\FOR {$k = 1, \dots, n-2$}
\STATE Let $C_k$ be the cluster induced by the node with rank $k$ in $R$ \label{alg:find_path:line:cluster}
\WHILE {$\rank_{T'}(\mrca_{T'}(C_k))>k$}
\STATE Update $T'$: Decrease $\rank_{T'}(\mrca_{T'}(C_k))$ by an $\rnni$ move \label{alg:findpath:line:move_set_down}
\STATE $p = p+T'$
\ENDWHILE
\ENDFOR
\RETURN $p$
\end{algorithmic}
\end{algorithm}

@@

\begin{proof}
We prove this theorem by backward induction on the length of a shortest caterpillar path $d_c(T,R) = l$ ($l$ is the induction parameter) between trees $T$ and $R$.
The backward induction basis follows by Lemma~\ref{lemma:caterpillar_dist=diameter}, as the maximum possible value for $l$ is $\frac{(n-1)(n-2)}{2}$.

Let $T$ and $R$ be caterpillar trees such that $d_c(T, R) = l < \frac{(n-1)(n-2)}{2}$.
Since $d_c(T, R) < \frac{(n-1)(n-2)}{2}$, there must be a step in $\csort(T, R)$ where the taxon that is moved up is not part of the cherry (of $T'$) at the beginning of this step.
Let $x$ be the first taxon that has this property and is considered at step $k$.
Then there must be a taxon $y$ that is lower neighbour of $x$ in the tree at the beginning of step $k$.
Since we assume that step $k$ is the first one that does not require the maximum number of $\nni$ moves, $y$ is lower neighbour of $x$ in $T$ as well.

Let $\hat T$ be a tree that equals $T$ but with $x$ and $y$ being exchanged.
Applying $\csort$ on $\hat T$ and $R$ results in almost the same path as the one computed by $\csort$ from $T$ to $R$, there is just one additional move in the beginning of step $k$ for swapping $x$ and $y$.
It follows directly that
\todo{No.
As $x$ and $y$ don't have to be neighbours in $T$.}
$d(T, \hat T) = 1$ and $d_c(\hat T,R) = d_c(T,R) + 1 = l + 1$.

Therefore, we can apply the induction hypothesis for $\hat T$ and $R$ and get $d(\hat T,R) = l+1$.
If it now were $d(T,R) < l$, it would be $d(\hat T,R) \leq d(T,R) + d(T,\hat T) < l + 1$ which contradicts the induction hypothesis.
We can conclude that $d(T,R) = l$, which completes the proof.
\end{proof}

>>>>

\todo{This format takes up too much space.
Why do we need END statements?}
\begin{algorithm}[H]
\caption{$\findpath$($T,R$)}
\label{alg:find_path}
\begin{algorithmic}[1]
\STATE $T' := T$, $p := [T']$
\FOR {$k = 1, \dots, n-2$}
  \STATE Let $C_k$ be the cluster induced by the node with rank $k$ in $R$ \label{alg:find_path:line:cluster}
  \WHILE {$\rank_{T'}(\mrca_{T'}(C_k))>k$}
    \STATE Update $T'$: Decrease $\rank_{T'}(\mrca_{T'}(C_k))$ by an $\rnni$ move \label{alg:findpath:line:move_set_down}
    \STATE $p = p+T'$
  \ENDWHILE
\ENDFOR
\RETURN $p$
\end{algorithmic}
\end{algorithm}

@@

\proof
Note that it is suffice to prove that $d_c(T, R) = d(T, R)$ for an arbitrary pair of caterpillar trees $T$ and $R$.
Lemma~\ref{lemma:caterpillar_dist=diameter} implies that if $T$ and $R$ are at the maximum possible distance, that is, $d_c(T, R) = \frac{(n-1)(n-2)}{2}$.
Denote the latter number by $D$.

Assume now that $d_c(T, R) = D - 1$.
If we apply $\csort(T, R)$ in this case, there must be a step in the algorithm where the taxon that is moved up is not part of the cherry (of $T'$) at the beginning of this step.
Let $x$ be the first taxon that has this property and is considered at step $k$.
Then there must be a taxon $y$ that is immediately preceding $x$ in the tree at the beginning of step $k$.
Since $k$ is the first such step, $y$ must be preceding $x$ in tree $T$ as well.
Consider a tree $\hat T$ that is obtained from $T$ by exchanging $x$ and $y$.
Note that the caterpillar path from $\hat T$ to $R$ computed by $\csort$ passes $T$ and coincides with the path $\csort(T, R)$ from there on.
Clearly, $d_c(\hat T, R) = D$.

Assuming $d_c(T, R) < D$, we can iterate the construction above to find a tree $\hat T$ such that $d_c(\hat T, R) = D$ and the caterpillar path from $\hat T$ to $R$ computed by $\csort$ passes $T$ and coincides with the path $\csort(T, R)$ from there on.
Lemma~\ref{lemma:caterpillar_dist=diameter} implies that $d(\hat T, R) = D$.

Assume that $d(T, R) < d_c(T, R)$ and let $p$ be a path from $T$ to $R$ of length $d(T, R)$.
Consider a path $q$ obtained by concatenating paths $\csort(\hat T, T)$ and $p$.
Note that $q$ is a path from $\hat T$ to $R$ which is shorter than $\csort(\hat T, R)$.
Indeed, the two paths coincide between $\hat T$ and $T$ and $q$ is shorter between $T$ and $R$.
Since $d_c(\hat T, R) = D$, the existence of $q$ is a contradiction to Lemma~\ref{lemma:caterpillar_dist=diameter}.
\endproof
